Project proposal: Explanations for unexplained effect size heterogeneity

### 2021-06-01 2020-05-17_proposal-unexplained-heterogeneity.rmd
Proposal v2

---
title: "Project proposal: Explanations for unexplained heterogeneity in meta-analysis"
author: "Anton Olsson Collentine"
date: '2021-05-17'
output: word_document
---

-	When studies fail to replicate heterogeneity is often raised as a potential explanation. 
-	This explanation is supported by most meta-analyses in psychology reporting large heterogeneity in the studied effect (Stanley, van Erp). Even in medicine, where heterogeneity estimates tend to be smaller (Ioannidis), the majority of meta-analyses appear to find heterogeneity (between 50-75%; Viechtbauer referencing Field; Higgins; Lipsey).
-	Some authors have argued that heterogeneity is inevitable in meta-analysis (Higgins), an argument which finds support in recent research demonstrating that substantial heterogeneity between studies can arise even from apparently minor differences in data processing (multiverse papers),  
-	Moreover, most preregistered multi-lab projects in psychology, where data collection and processing is identical across studies, seem to find little or no heterogeneity (Olsson-Collentine et al.)
-	It seems like in most situations (the exception being multi-lab projects or meta-analyses with very strict inclusion criteria) we should expect meta-analyses to have heterogeneity 
-	However, heterogeneity estimates in meta-analysis are almost invariably discussed as representing differences in true effect sizes between included studies (e.g., refs)
-	That is, arising from moderator effects
-	However, heterogeneity estimates in meta-analysis may also arise from statistical artefacts.
-	These may derive from, for example, the estimation procedure (von Hippel), differential measurement error  (Olsson-Collentine et al), publication bias (Augusteijn), and even arbitrary decisions in the research process (multiverse papers)
-	Fundamentally, anything that affects effect size estimates may also affect statistical heterogeneity estimates.
-	As such, researchers should make a concerted effort to rule out artefactual explanations of statistical heterogeneity before examining moderators
-	In this paper, we describe several explanations for unexplained heterogeneity in meta-analysis, examine their general effect on heterogeneity estimates and how to rule them out. 
-	We first provide a short definition of heterogeneity and how it is quantified.

#### What is heterogeneity 
-	It may be useful to distinguish between clinical heterogeneity (differences in contextual factors between studies) and statistical heterogeneity (a statistical estimate of the sensitivity to differences in contextual factors between studies).  
-	A statistical definition of heterogeneity may make it more clear why ‘heterogeneity’ cannot be assumed to directly represent the differences in true effects between included studies.
-	We can statistically define heterogeneity as: the residual variance between effect sizes after taking into account sampling error. 
-	Mention something about estimation, but probably don’t go into details (provide references).
-	Mention different estimators (I2, Q, Tau2), but discuss them further when talking about explanations

## Methods

#### Potential empirical aspects to add
a)	Strictness of inclusions medicine vs psychology
    - How many papers do meta-analysts exclude from their initial search?
      - Check PRISMA flowcharts in meta-analyses.
    - And/or compare the number of inclusion criteria
b)	Literature review of different QRPs that affect heterogeneity (“poor methodological quality”/ “small study effects”)
c)	Moderators (% explained variance vs. power)
    - Perhaps we can just do a smallish review, e.g., look at Robbie’s meta-meta data?
d) Check how often I2 used over tau (might already be some paper who has done this)
e) Check how often sample sizes in primary studies are large enought to lead to substantial I2 heterogeneity
f) Check how often reliability is available in primary studies (or refer to e.g., Flake et al)
g)	Use the meta-multiverse data to demonstrate the effect of arbitrary decisions on heterogeneity

## Explanations for heterogeneity
-	Here we first list potential artefactual explanations and how they might be examined. We then shortly discuss examining potential moderators of true heterogeneity, but primarily refer the reader to the several excellent papers available on this topic.

#### Explanation 1: estimation artefacts
a) I2 -> heterogeneity increases with primary study sample sizes
    - Rule out by using tau instead
b) Central chi-square distribution (Q->I2/tau) can lead to heterogeneity (von Hippel)
    - If heterogeneity is large enough, not a sufficient explanation = can rule out

#### Explanation 2: measurement artefacts
-	Differential measurement error
    - Can examine how large the effect would be on estimates if we know reliability in all primary studies? = rule out as sufficient explanation if heterogeneity too large
-	Range restrictions? (Schmidt & Hunter, 2015; cited in Olsson-Collentine et al.)
-	“poor methodological quality”

#### Explanation 3: arbitrary data processing and model decisions
-	Multiverse papers
    - Can never rule out as explanation (maybe with access to raw data?)

#### Explanation 4: QRPs
-	Publication bias (Hilde)
    - Can try to rule out with inclusion of grey literature, publication bias tests
-	Selective reporting (including p-hacking) ((“small study bias”)
    - Different forms of p-hacking? 
    - Can rule out only by checking preregistration

#### Explanation 5: systematic variance (moderators)
-	Short explanation

## Summary & Conclusion
-	There are many potential explanations for why the purportedly same true effect size differs between studies and results in heterogeneity in meta-analysis.
-	Before examining moderators as an explanation, researchers should make a concerted effort to rule out artefactual explanations such as those discussed in this paper.

>>>>

---
title: "Project proposal: Explanations for unexplained effect size heterogeneity"
author: "Anton Olsson Collentine"
date: '2021-06-01'
output: word_document
---

## Short summary

-	Theoretical (mainly) paper consisting of introduction + 7 explanations + empirical section + conclusion
-	Sections are:
    - Introduction (1-2 pages)
        - Content: Comparable studies can have different outcome (heterogeneity), we present 7 explanations, argue that artefactual explanations should be examined before moderators, and summarize the content of the paper.
    1) Model and sampling error (1.5 – 2 pages)
        - Present meta-analytic heterogeneity model
        - Explain that it accounts for sampling error
        - Discuss estimation artefact that can increase heterogeneity estimate (I2 vs. tau, central chi-square distribution of Q)
    2) Multiverse effect (2 pages)
        - Discuss based on multi-analyst papers
        - Illustrate using data from meta-multiverse project
        - Highlight that even though multiverse does not affect average estimates in absence of selective reporting, it can affect heterogeneity estimates
    3) Publication bias (max 1 page)
        - Literature review: Hilde’s paper + others
    4) QRPs (minimum 2 pages with illustrations)
        - should usually decreases heterogeneity, assuming people select for significance
        - Literature review on different types of QRPs (talk to Esther)
    5) “Incompetence”/ “poor methodological quality” (1/3 of a page)
        - Common critique when replications ‘fail’
        - This critique is not detailed enough as a critique
        - If an “incompetent” researcher uses a less effective intervention, they should be seen as a design moderator (but must first be better specified)
        - Cochrane recommends talking about ‘risk of bias’ rather than ‘poor methodological quality’
    6) Measurement artefacts (1 – 2 pages)
        - Differential measurement error leads to heterogeneity (Flake et al., examined availability of reliability estimates in primary studies)
        - Range restrictions? Other measurement factors? (see Schmidt & Hunter, 2015)
    7) Moderators (2 pages)
        - Short discussion, extensively described in other papers
        - Types of moderators (contextual factors): settings, sample population, treatment and measurement variables
        - Meta-regression, power, IPD
    - Methods & results (2  pages)
        - Description of empirical data collection
        - Discuss findings in the context of the section(s) 1- 7 that they affect
    - Conclusion (1/2 page)
-	Total: ~15 pages

## Detailed summary
Below I present a detailed summary of

a) the introduction: for a better description of the proposed angle of the paper. 
b) Methods & results section: my proposals for the empirical data collection. Needs to be discussed to set the scope of the paper.

### Introduction

**Comparable studies always have some design heterogeneity, whether they differ in the studies’ settings, their sample populations, the measures they use, or their treatment variables (e.g., Campbell & Stanley, 2015).** In turn, effect size estimates from comparable studies always have some statistical heterogeneity, one potential explanation of which is the studies’ design heterogeneity (i.e., the effect size estimates arising from different true effects). Other explanations for differences in statistical estimates from comparable studies include sampling error, multiverse variability, publication bias, Questionable Research Practices such as selective reporting (including p-hacking), ‘poor methodological quality’ or supposed incompetence of experimenters, and measurement artefacts.

**It can be challenging to disentangle the reasons for why effect size estimates differ.** From a theoretical perspective we are most interested in finding out to what extent these differences arise from design heterogeneity, to which end other explanations are nuisances or artefacts we must first control for. Perhaps this explains why heterogeneity estimates in meta-analysis are almost invariably discussed as representing differences in true effect sizes between included studies, ignoring the many alternative explanations.

**Generally, when comparable studies show different results, there is an undue focus on design heterogeneity as an explanation for these differences.** Design heterogeneity is often one of the first explanations for any ‘failure’ to replicate a study (by the reader’s preferred definition), of which the ‘replication crisis’ in psychology and other domains continue to provide many examples. However, artefactual explanations for statistical heterogeneity by necessity should be partitioned out first before considering design heterogeneity as an explanation. In addition, design heterogeneity is often an unlikely explanation for statistical heterogeneity in at least in cognitive and social psychology (Olsson-Collentine et al, 2020). Fundamentally, anything that affects effect size estimates may also affect statistical heterogeneity estimates.As such, researchers should make a concerted effort to rule out artefactual explanations of statistical heterogeneity before examining design heterogeneity as an explanation.

**The statistical results from comparable studies and their heterogeneity are usually examined through meta-analysis, and this is also the tool we use throughout this paper to discuss explanations for statistical heterogeneity.** Examining statistical heterogeneity and moderators thereof is usually considered one of the primary purposes of meta-analysis. To enhance the discussion of statistical heterogeneity , we will also collect empirical data relating to heterogeneity estimates in meta-analysis. We first describe and discuss all 7 explanations for statistical heterogeneity, then present a short methods section and discuss relevant sections in the context of our findings, before ending the paper with a summary and conclusion.

### Methods & Results
- This section has a large impact on the scope of the project and needs discussion on what we should actually do. 
- I first describe 3 research questions we might try to answer by collecting data, and then make different suggestions to limit the scope.

#### Research questions and empirical data collection:
  1) To what extent does meta-analysts’ usage of I2 instead of tau inflate heterogeneity estimates?
        - Extract I2 and harmonic mean
  2) To what extent does the smaller average number of studies in medical meta-analyses (compared to in psychology) represent stricter inclusion criteria? (as opposed to smaller population of studies)
        - Extract proportion excluded studies from flowcharts
        - Extract number of exclusion criteria 
        - Extract type of exclusion criteria (settings, sample population, measurement or treatment variables, risk of bias)
  3) To what extent and which type of moderators explains statistical heterogeneity in meta-analyses?
        - Do the meta-analyses examine moderators? {yes/no}
        - What % of statistical heterogeneity do moderators explain?
        - How many moderators are examined, and what type? 
            - 	Settings, sample population, measurement and treatment variables, risk of bias
            - 	What % does each type explain?
        - What is the power of the meta-analysis to find significant moderators?

#### Scoping (suggestions for different levels of empirical data collection)
- The moderator part above might be an entire paper by itself and we may not want to do everything. Below 5 suggestions for limiting the scope of data collection in this paper. Other suggestions?
- My inclination would be towards suggestion 3/4 below, which would leave most of the analysis of moderators in meta-analysis to be part of a future project.

Suggestions:

  1) Extract:
        - I2 + harmonic mean
        - Proportion excluded studies (medicine vs. psychology)
  2) Extract:
        - I2 + harmonic mean
        - Proportion excluded studies (medicine vs. psychology)
        - PLUS: Number and type of exclusion criteria
  3) Extract:
        - Proportion excluded studies (medicine vs. psychology)
        - Number and type of exclusion criteria
        - PLUS: How many moderators examined and what type [e.g., psychology might have fewer exclusion criteria, but more moderators]
  4) Extract:
        - Proportion excluded studies (medicine vs. psychology)
        - Number and type of exclusion criteria
        - How many moderators examined and what type [e.g., psychology might have fewer exclusion criteria, but more moderators]
        - PLUS: I2 + harmonic mean
  5) Extract:
        - Everything, including all the moderator stuff

#### Sample 
What would the sample look like?

  -	A random sample of meta-analyses from medicine and psychology
      - Already collected sample(s) from some project? Our inclusion/exclusion criteria? Sample size?
